Automatic word stress annotation of Russian running text (max 8 pages)

### 2015-01-19 rusaccent-nodalida-2015/rusaccent.tex
more
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58450 72bbbca6-d526-0410-a7d9-f06f51895060

\usepackage{linguex}

@@

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of tools we developed for automatically annotating 
  word stress in Russian running text. Given an adequate lexicon with specified stress, the
  primary obstacle for correct stress placement is disambiguating homographic forms. Using a
  Constraint Grammar, we achieve a X\% improvement over the baseline undisambiguated performance.
\end{abstract}

@@

Russian word stress is variable. The inflecting word classes include complex patterns of
shifting stress, and a lexeme's stress pattern cannot be predicted from surface forms.
Although lexical stress and its attendant vowel reduction are a prominent feature of
spoken Russian, standard written Russian does not typically mark word stress.\footnote{
Texts intended for native speakers sometimes mark stress on words that cannot be 
disambiguated through context, although this is somewhat rare.} Without information about lexical 
stress position, 
correctly converting written Russian text to speech is impossible. This is true both for humans
(e.g. foreign language students) and computers (e.g. text-to-speech).

@@

\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \rus{т\'{е}ла} \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' \label{ex:bodySGGEN}
\b. \rus{тел\'{а}} \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' \label{ex:bodyPLNOM}

The remaining two types of stress ambiguity occur between lexemes. \emph{Morphosyntactically
congruent} stress ambiguity occurs between homographs that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in Table~\ref{table:MScongruent}.

@@

So far, the implicit target application the few studies related to automatic stress placement
in Russian has been text-to-speech and automatic speech recognition. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

So far, the approaches discussed are
dependent on the availability of readings from the FST. Two final approaches address unknown tokens.
The {\small {\tt guessSyll}} and {\small {\tt guessFreqSyll}} approaches are identical to 
{\small {\tt guess}} and {\small {\tt guessFreq}}, respectively,
except that for unknown tokens, they places the stress near the end of the stem.\footnote{Determining 
which vowel is stem-final
is itself a difficult task. Our algorithm merely places the stress on the last vowel that is 
followed by a consonant.} \rr{Explain why stem-final is a good guess (and cite someone?)}

For our baseline, we take the output of our morphological analyzer (without the Constraint 
Grammar) in combination with the {\small {\tt bare}}, {\small {\tt safe}}, 
{\small {\tt guess}} and {\small {\tt guessFreq}} approaches.

We also compare our outcomes with the RussianGram\footnote{http://russiangram.com/} plugin 
for the Google Chrome web browser. RussianGram is not open-source, so we
can only guess what technologies support the service. In any case, it provides a
meaningful reference point for the success of each of the methods described above.

@@

\begin{itemize}
\item Examine errors and categorize/refute them.
\end{itemize}

@@

One of the main points of this paper is to highlight the importance of 
syntactic context in the Russian wordstress placement task. \cite{hall_sproat_russian_2013} 
achieve 98.7\% accuracy on stress placement of individual wordforms. To 
understand this result, it should be remembered that Russian has a considerable
number of words that are ambiguous with regard to stress. Based on the surface 
forms in our FST (which is based on the same dictionary used for 
\cite{hall_sproat_russian_2013}), we calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions.} 
Compare this with the finding of our experiment, that in running 
text, at least 7.5\% of the tokens are stress-ambiguous. This is at least partly 
because many of 
the lexemes with shifting stress patterns are high-frequency. Any system that
cannot leverage syntactic context to disambiguate these wordforms cannot 
achieve accuracy above 92.5\% without making an higher number of errors.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.
The experiment described in this paper demonstrates that a Constraint Grammar
can effectively improve the accuracy of a stress placement system without 
significantly raising the error rate. Our Russian Constraint Grammar is
under continual development, so we expect higher accuracy in the future.

The experiment was limited by properties of the gold standard corpus, including
its size, genre distribution, and quality. Ideally, a gold standard corpus for Russian wordstress should mark 
whether monosyllabic words are stressed, and of course its annotations should
be accurate. Our gold corpus represents a broad variety of text genres, which 
makes our results more generalizable, but a larger corpus would allow for
evaluating each genre individually. For example, the vast majority of words with
shifting stress are of Slavic origins, so we expect a genre such as technical
writing to have a lower proportion of words with stress ambiguity, since it
uses a more modern lexicon that has not inherited the idiosyncrasies that lead
to wordstress ambiguity.

In addition to genre, it seems likely that text
complexity affects the difficulty of the stress placement task, so subcorpora at
various reading levels would shed light on those interactions as well.

The syllable-guessing algorithm used in this experiment was overly simplistic,
and it was not surprising that it was

>>>>

\usepackage{linguex}
\usepackage{needspace}

@@

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of finite-state tools we developed for 
  automatically annotating word stress in Russian running text. To our knowledge,
  this is the first study to empirically evaluate the results of this task. 
  Given an adequate 
  lexicon with specified stress, the primary obstacle for correct stress 
  placement is disambiguating homographic wordforms. Using a Constraint Grammar, 
  we achieve 93.21\% accuracy, a 3.14\% improvement over the baseline
  undisambiguated performance. 
  Combining our Constraint Grammar with frequency data and a simple
  algorithm for unknown word stress position guessing, we achieve 96.15\%
  accuracy. These results highlight the need morphosyntactic disambiguation
  in the wordstress placement task for Russian, and set a standard for future
  research on this task.
\end{abstract}

@@

Russian word stress is variable and phonemic. The inflecting word classes include complex 
patterns of shifting stress, and a lexeme's stress pattern cannot be predicted 
from surface forms. Lexical stress and its attendant vowel reduction are 
a prominent feature of spoken Russian, and the incorrect placement of stress
can render speech incomprehensible. However, standard written Russian does not 
typically mark word stress.\footnote{Texts intended for native speakers sometimes 
mark stress on words that cannot be disambiguated through context, although this 
is rare.} Without information about lexical stress position, 
correctly converting written Russian text to speech is impossible. This is true 
both for humans (e.g. foreign language students) and computers (e.g. text-to-
speech).

@@

\needspace{4\baselineskip}
\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \rus{т\'{е}ла} \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' 
    \label{ex:bodySGGEN}
\b. \rus{тел\'{а}} \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' 
    \label{ex:bodyPLNOM}

The remaining two types of stress ambiguity occur between lexemes. 
\emph{Morphosyntactically congruent} stress ambiguity occurs between homographs 
that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in
Table~\ref{table:MScongruent}. This kind of stress ambiguity is relatively rare.
Resolving this ambiguity would require the use of  Word Sense Disambiguation, which is
outside the scope of this paper.

@@

So far, the implicit target application of the few studies related to automatic stress placement
in Russian has been text-to-speech and automatic speech recognition. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

So far, the approaches discussed are
dependent on the availability of readings from the FST. Two final approaches 
address unknown tokens.
The {\small {\tt guessSyll}} and {\small {\tt guessFreqSyll}} approaches are 
identical to {\small {\tt guess}} and {\small {\tt guessFreq}}, respectively,
except that the {\small {\tt Syll}} approaches attempt to place stress on
unknown tokens (i.e. tokens not found in the FST). Empirical research with
Russian native speakers reading nonce words in running text concludes that
the default stress position in Russian is the last syllable of the stem, or
in other words, the last vowel that is not part of an inflectional ending 
\cite{Crosswhite.ea-03}. Therefore, the {\small {\tt Syll}} approaches 
place the stress near the end of the stem. Determining 
which vowel is stem-final
is itself a difficult task. Our algorithm merely places the stress on the last 
vowel that is followed by a consonant. Although more sophisticated algorithms
are necessary to accurately locate the stem boundary, this is not the central
focus of this study.

For our baseline, we take the output of our morphological analyzer (without the 
Constraint Grammar) in combination with the {\small {\tt bare}}, 
{\small {\tt safe}}, {\small {\tt guess}} and {\small {\tt guessFreq}} 
approaches. We also compare our outcomes with the 
RussianGram\footnote{http://russiangram.com/} plugin 
for the Google Chrome web browser. RussianGram is not open-source, so we
can only guess what technologies support the service. In any case, it provides a
meaningful reference point for the success of each of the methods described 
above.

@@

\rr{Examine errors and categorize/refute them.}

@@

One of the main points of this paper is to highlight the importance of 
syntactic context in the Russian wordstress placement task. 
\newcite{hall_sproat_russian_2013} achieve 98.7\% accuracy on stress placement of 
individual wordforms. To 
understand this result, it should be remembered that Russian has a considerable
number of words that are ambiguous with regard to stress. Based on the surface 
forms in our FST (which is based on the same dictionary used for 
\cite{hall_sproat_russian_2013}\rr{need to cite this without parentheses}), we 
calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions.} 
Compare this with the finding of our experiment, that in running 
text, at least 7.5\% of the tokens are stress-ambiguous. This is at least partly 
because many of 
the lexemes with shifting stress patterns are high-frequency. Any system that
cannot leverage syntactic context to disambiguate these wordforms cannot 
achieve accuracy above 92.5\% without making an higher number of errors.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.
The experiment described in this paper demonstrates that a Constraint Grammar
can effectively improve the accuracy of a stress placement system without 
significantly raising the error rate. Our Russian Constraint Grammar is
under continual development, so we expect higher accuracy in the future.

We are unaware of any other studies of Russian wordstress placement in running
text. The results of our experiment are promising, but many questions remain
unanswered.
The experiment was limited by properties of the gold standard corpus, including
its size, genre distribution, and quality. Ideally, a gold standard corpus for 
Russian wordstress should mark 
whether monosyllabic words are stressed, and of course its annotations should
be accurate. Our gold corpus represents a broad variety of text genres, which 
makes our results more generalizable, but a larger corpus would allow for
evaluating each genre individually. For example, the vast majority of words with
shifting stress are of Slavic origins, so we expect a genre such as technical
writing to have a lower proportion of words with stress ambiguity, since it
uses a more modern lexicon that has not inherited the idiosyncrasies that lead
to wordstress ambiguity.

In addition to genre, it is also likely that text
complexity affects the difficulty of the stress placement task. The distribution
of different kinds of syntactic constructions vary with text complexity 
\cite{Vajjala.Meurers-12}, and
we expect that the effectiveness of the Constraint Grammar will be affected by
those differences. used in would shed light on those interactions as well.

The syllable-guessing algorithm {\small {\tt Syll}} used in this experiment was 
overly simplistic, and it was not surprising that it was only moderately 
successful. More rigorous approaches have been suggested in other studies.
One approach -- exemplified by \newcite{Church-85}, 
\newcite{Williams-87} and \newcite{Xomicevic_2008} -- is to 
attempt to parse the unknown token by matching known prefixes and suffixes.
Another approach -- exemplified by Pearson.ea-00, Webster-04, Dou.ea-09, and 
\newcite{hall_sproat_russian_2013} -- is based on machine-learning. For unknown 
words, \newcite{hall_sproat_russian_2013} achieve an accuracy of 83.9\%.
It should be noted that their model was trained on a full list of Russian words,
which is not representative of the words that would be unknown to a system like
ours. Most of the complicated wordstress patterns are a closed 
class.\footnote{The growing number of masculine nouns with shifting stress are 
the exception to this generalization.} Future machine-learning work on Russian 
wordstress placement in running text should limit the training data to word
classes that are likely to be similar to unknown tokens, such as those with
productive derivational affixes.
